Density-dependent selection and the limits of relative fitness

### 2017-12-05 manuscript.tex
l

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In continuous time, we have the canonical selection equation \citep[pp. 26]{crow_1970}
\begin{equation}
\frac{d p_i}{dt}=(r_i-\overline{r})p_i, \label{eq:canonical}
\end{equation}
where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$. If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as $\frac{d p_i}{dt}=s p_i(1-p_i)$, where the constant selection coefficient $s$ is the difference in $r$ between types; the corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. This is exemplified by the Wright-Fisher model. Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others. Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). Thus, $N$ changes during, and as a result of, the adaptive sweeps described by Eq.~\eqref{eq:canonical}. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

@@

%But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.

@@

\section*{The limits of relative fitness}

We are now in a position to analyze the validity of relative fitness models more rigorously. To do so, we evaluate some objections to Eq.~\eqref{eq:canonical} in a crowded poulation using our density-dependent lottery model as a benchmark to better understand the population ecological implications.

We first set aside the most obvious objection: the selection coefficient will often be density-dependent because crowded and uncrowded conditions are so different. This objection cannot be ignored if $N$ changes substantially during a sweep. We return to this in the Discussion. For now we restrict our attention to the case that $N$ is at its equilibrium value.

With $N$ at equilibrium, the presence of density-dependent selection is beside the point. As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps. For $b$ and $d$ sweeps, $s$ is also exactly consant, because the overall regulation of population density applies equally to all types \citep{prout_1980}. To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
\begin{equation}
\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
\end{equation}
Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$. This applies uniformly to all types and therefore does not induce density-dependence in $b$ or $d$ selective advantages. Thus, $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly. That is, in our lottery model density will adjust itself as a result of selection, selection is density-dependent, and yet Eq.~\eqref{eq:canonical} applies exactly for pure $b$, $c$ and $d$ sweeps.

As noted in the Introduction, widely used models of population growth predict that selection in crowded populations will intrinsically affect population density. In the previous section we revisited MacArthur's argument and saw that there is no reason to expect this to be true in general, and indeed the competitive ability trait $c$ in our lottery model does not have this property.

>>>>

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In continuous time, we have the canonical selection equation $\frac{d p_i}{dt}=(r_i-\overline{r})p_i$, where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as 
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. This is exemplified by the Wright-Fisher model. Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others. Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness''). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

@@

@@

\section*{Type-specific responses to density and the strength of selection}

We are now in a position to analyze the validity of relative fitness models more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp. 29]{crow_1970}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even in crowded populations where $N$ has reached equilibrium, because adaptive sweeps in density-dependent traits then induce a change in $N$ and also change themselves in response to $N$. To quantify how serious this objection is, we need to account for how much $N$ changes over a sweep. In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N=b_i/\delta_i$, and so the selection coefficient increases from $s_{\rm initial}$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a sweep (Fig.~\ref{fig:strengthofselection}). Thus, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on the density-dependent mortality rate, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection.

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-selection''-type sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding. The population is in demographic equilibrium at the start of the sweep.}
\end{figure}

Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:Ndependentsweep} with our density-dependent lottery. As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps. For $b$ and $d$ sweeps, $s$ is also exactly consant, because the overall regulation of population density applies equally to all types \citep{prout_1980}. To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
\begin{equation}
\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
\end{equation}
Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$. This applies uniformly to all types and therefore does not induce density-dependence in selective advantages related to $b$ or $d$. Note that our lottery model accounts for two key aspects of the interaction between selection and demography: selection is density-dependent, and density will change as a result of selection on absolute fitness traits. And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.

The reason that our lottery model behaves so differently from Eq.~\eqref{eq:simplebirthdeath} is that each type can respond differently to density via $c$ without affecting the regulation of population density. That is, our lottery model separates density-dependent selection from ``selection-dependent density'' \citep{prout_1980}, two distinct effects that are conflated in simple growth models with linearly-density-dependent per-capita growth rates. If types in the lottery model differ in $c$ as well as $b$ or $d$, then the lottery model will exhibit similar complications as Fig.~\ref{fig:strengthofselection}.
